Degrees of Difference: Analyzing the Impact of Education on Earnings in Canada

### 2024-04-19 paper/paper.qmd
updated paper structure and scripts, added sketches

---
title: "Degrees of Difference: Analyzing the Impact of Education on Earnings in Canada"
author: 
  - Yuanyi (Leo) Liu
thanks: "Code and data are available at: https://github.com/leoyliu/Analyzing-the-Impact-of-Education-on-Earnings-in-Canada"
date: today
date-format: long
abstract: "In this paper, we explore the connection between education and hourly wages in Canada, focusing on data from the year 2000. Our findings indicate a clear trend: higher educational levels correlate with increased average hourly wages across all age categories. This research highlights the importance of education in determining earning potential and suggests that investment in education could have long-term economic benefits. The study provides evidence for policy implications regarding educational incentives and workforce development strategies in Canada."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("here")
# install.packages("mapdata")
# install.packages("ggplot2")
# install.packages("haven")
# install.packages("kableExtra")

library(tidyverse)
library(dplyr)
library(here)
library(mapdata)
library(ggplot2)
library(haven)
library(kableExtra)
```

@@

The relationship between education and earnings is a well-established topic of interest within the field of labor economics, providing insights into the broader socio-economic fabric of a country. As Canada's economy continues to diversify and specialize in various sectors, the value of education in this landscape remains a critical question for policymakers and the public. While previous studies have explored this link, there is an ongoing need to update and deepen our understanding of how this dynamic plays out across different age groups in the modern economy.

This paper aims to fill the knowledge gap by analyzing the impact of education levels on the average hourly wage rate in Canada, with a specific focus on three age cohorts: young entrants (15-24 years), established workers (25-54 years), and senior professionals (55 years and over). The analysis begins in the year 2000, a period of significant technological advancement and economic change, offering a contemporary perspective on the education-wage relationship. By dissecting the data across these age groups, the study provides nuanced insights into the economic returns of education at various career stages.

Our research methodology involved a detailed statistical analysis of wage data, stratified by education level and age group, to uncover patterns and variations in earnings. The findings confirm the hypothesis that higher education correlates with increased wages, with the effect becoming more significant as workers age. This not only reinforces the case for educational investment but also suggests that such investment pays dividends throughout an individual’s career, not just in the early years.

In this study, our primary focus is to estimate the effect of education on hourly wages across different age groups in Canada. Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level, while controlling for age. By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums across various stages of a typical Canadian worker’s career span.

@@

This section aims to offer an insightful understanding of the dataset utilized in our analysis. The dataset captures the birth rates across various demographics in the United States from 1980 to 2020. The data provide a broader context, allowing an analysis of trends over time and across different economic cycles, including the period of the Great Recession.

## Source and Methodology
Data on birth rates for women aged 15-44 were sourced from National Vital Statistics Reports covering 2015, 2019, and 2020 [@cite2015birth; @cite2019birth; @cite2020birth], compiled from birth certificates across the US and the District of Columbia. @citeoriginal provided aggregated birth data for six cohorts over five-year spans, derived from the NBER Natality Database [@citeNBER] and NCHS microdata [@citeNCHS] for 1980-2019. In this study, we simply used the replication data package provided by @citeoriginal for analysis.

While there were alternative datasets available from other public and private sources, this dataset was chosen due to its comprehensive coverage, reliability, and the level of detail it offers. It includes data from the National Center for Health Statistics [@citeNCHS] and the U.S. Census Bureau, which are both reputable sources of demographic data.

The data was processed and cleaned using R[@citeR], a powerful statistical programming language. For key operations, please refer to the [@sec-appendix].

@@

To better understand the data, a summary table was developed to provide a detailed description of each variable, explaining its relevance and how it contributes to our understanding of the topic. Our focus is not only on the direct measures of birth rates but also on a range of demographic and economic variables that offer a comprehensive view of the factors influencing fertility decisions during the recessionary period.

@@

The measurement of birth rates in our study involves a detailed analysis of the annual number of live births per 1,000 women in various age groups, ranging from 15 to 44 years. These rates were extracted from datasets provided by the National Center for Health Statistics (NCHS) [@citeNCHS], ensuring accuracy and reliability in capturing demographic trends across the United States from 1980 to 2020.

For each age group, birth rates were calculated by dividing the total number of live births by the population of women in that age range, then multiplying by 1,000 to standardize the measure. This approach allows for comparison across different demographics and time periods, providing a clear picture of how birth rates have shifted, particularly in response to economic conditions such as the Great Recession.

In addition to age-specific birth rates, state-level data were also analyzed to identify geographic variations in birth rate trends. This involved mapping birth rates against state populations, taking into consideration the exclusion of data from states like Alaska and Hawaii where mapping constraints exist. Further details on this aspect of our study are discussed in @sec-result.

@@

The goal of our modelling strategy is twofold. Firstly,...

@@

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

>>>>

---
title: "Degrees of Difference: Analyzing the Impact of Education on Earnings in Canada"
author: 
  - Yuanyi (Leo) Liu
thanks: "Code and data are available at: https://github.com/leoyliu/Analyzing-the-Impact-of-Education-on-Earnings-in-Canada"
date: today
date-format: long
abstract: "In this paper, we explore the connection between education and hourly wages in Canada, focusing on data from the year 2000. Our findings indicate a clear trend: higher educational levels correlate with increased average hourly wages for individuals aged 25 to 54. This research highlights the importance of education in determining earning potential and suggests that investment in education could have long-term economic benefits. The study provides evidence for policy implications regarding educational incentives and workforce development strategies in Canada."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("here")
# install.packages("mapdata")
# install.packages("ggplot2")
# install.packages("haven")
# install.packages("kableExtra")
# install.packages('tinytex')
# tinytex::install_tinytex()
# to uninstall TinyTeX, run tinytex::uninstall_tinytex()

library(tidyverse)
library(dplyr)
library(here)
library(mapdata)
library(ggplot2)
library(haven)
library(kableExtra)
library(tinytex)
```

@@

The relationship between education and earnings is a well-established topic of interest within the field of labor economics, providing insights into the broader socio-economic fabric of a country. As Canada's economy continues to diversify and specialize in various sectors, the value of education in this landscape remains a critical question for policymakers and the public. While previous studies have explored this link, there is an ongoing need to update and deepen our understanding of how this dynamic plays out in the modern economy.

This paper narrows the focus to the Canadian labor market, where the interplay between educational achievement and wages within the core working age group, those aged 25-54, offers a mirror to the efficacy and value of educational advancements in the workplace. This demographic is selected for its relevance to the labor force and policy implications, excluding younger individuals still likely in the education system and older individuals, who may present outlier educational experiences that do not align with the central workforce.

This investigation seeks to address a gap in the current literature by providing a targeted analysis of the 25-54 age group in Canada since the year 2000, a period marked by rapid technological change and economic evolution. By applying a linear model to wage data categorized by education level, this study uncovers a positive correlation between educational attainment and hourly wages, suggesting that higher education can be linked to improved wage outcomes in this key demographic.

In this study, our primary focus is to estimate the effect of education on hourly wages in Canada. Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level. By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums across various stages of a typical Canadian worker’s career span.

@@

This section aims to offer an insightful understanding of the dataset utilized in our analysis, which serves as the foundation for our examination of the relationship between education and average hourly wage rates in Canada, specifically within the 25-54 age demographic.

## Source and Methodology

This study utilizes a dataset sourced from the Open Government Portal of Canada [], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce. The dataset covers the period from the year 2000 onwards, providing a longitudinal view of wage trends in relation to educational background.

Alternative datasets, such as those from Statistics Canada’s Labour Force Survey, were considered but were not selected due to their less detailed categorization of education levels and their broader focus on employment without specific wage breakdowns by education.

The data was processed and cleaned using R[@citeR], a powerful statistical programming language. The selection of this dataset was motivated by its direct relevance to the research question and its regular updates, ensuring that the data remains relevant for observing current trends. Initial data processing involved filtering the dataset to focus solely on entries categorized under 'Canada' in geography, eliminating data irrelevant to the national focus of this research. This step was crucial to maintaining the clarity and relevance of the analysis. For key operations, please refer to the [@sec-appendix].

@@

To better understand data, key variables extracted for this study include the year of data collection, education level, age group, and the average hourly wage rate. Notably, we have streamlined the 'Education level' from a broad range of categories into a numeric variable that aligns with ascending educational attainment, facilitating a quantitative analysis of its impact on wages. The age group has been limited to the 25-54 years range to focus on the most economically active segment of the population, avoiding potential outliers from younger individuals with less work experience and older individuals whose advanced education, such as doctoral degrees, could distort the analysis. Wage rates are presented in Canadian dollars per hour, reflecting pre-tax earnings, inclusive of tips, commissions, and bonuses.

```{r}
#| label: tbl-display-clean-data
#| tbl-cap: First Ten Rows of Wage Rates Spanning from 2000 to 2019 For Cleaned Data
#| echo: false

cleaned_data <- read.csv(here::here("data/analysis_data/analysis_data.csv"))

cleaned_data |>
  select(Year, Education.level, Age.group, Avg.hourly.wage.rate) |>
  slice(1:10) |>
  kable(
    col.names = c("Year", "Education Level", "Age", "Hourly Wage"),
    digits = 1,
    booktabs = TRUE
  )
```
@tbl-display-clean-data, created with `kableExtra` [@citekableExtra], outlines the first ten records of our dataset, featuring average hourly wages for Canadians aged 25-54 across varying levels of education from the years 2000 to 2019. This snapshot reveals the variable structure of our analysis, which includes the `Year` of wage data, the defined `Education level` ranging from "0 - 8 years" to "Above bachelor’s degree," and the consistent `Age group` focus. It provides an early indication of the ascending trend in wages with higher educational attainment, setting the stage for our deeper investigation into the economic value of education within Canada's workforce.

To transition from a tabular overview to a more graphical representation, we will visualize the data, which will allow us to spot patterns and trends that are not immediately apparent in numerical form.

```{r fig.width=7, fig.height=3}
#| label: fig-display-clean-data
#| fig-cap: Visualization of Wage Rate from 2000 to 2019
#| echo: false

ggplot(cleaned_data, aes(x = Year, y = Avg.hourly.wage.rate, color = Education.level)) +
  geom_point(alpha = 0.7, size = 2) + 
  labs(
    x = "Year",
    y = "Average Hourly Wage Rate (CAD)",
    color = "Education Level"
  ) +
  theme_minimal(base_size = 10) +  # Reduced base_size for smaller overall text
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),  # Smaller font for x-axis labels
    axis.text.y = element_text(size = 8),  # Smaller font for y-axis labels
    legend.position = "right",
    legend.text = element_text(size = 8)  # Smaller font for legend text
  ) +
  scale_color_brewer(palette = "Set1")  # Keep the color palette for distinction
```
@fig-display-clean-data illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54. Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.

@@

The measurement of variables within this dataset was handled to ensure accuracy and relevance. The 'Average Hourly Wage Rate' is calculated based on actual wages reported by employers, providing an objective measure of income. This data is then vetted and standardized by statistical agencies to ensure it reflects accurate and fair representations of wages across different demographics.

For educational attainment, data entry occurs as individuals enter the workforce or update their qualifications, with categorizations reflecting the highest level of formal education completed. This data is collected through surveys and employment records, often as part of broader demographic data collection efforts by governmental agencies.

For an in-depth technical breakdown of the statistical analyses and additional visualizations, readers are directed to the appendices, where these aspects are discussed in detail. This ensures that the main text remains accessible while still providing resources for more technically inclined readers.

Correlations between the variables will be examined at @sec-model, providing insights into how education levels are associated with hourly wages. By understanding these relationships, we can better infer the potential impact of education on earnings within the Canadian economy.

@@

The objective of our modeling approach is to quantify the relationship between educational attainment and hourly wages. Our analysis employs a Bayesian framework to assess how changes in the level of education correlate with variations in wage rates among Canadian workers aged 25-54.

@@

Define $y_i$ represent the average hourly wage rate for the $i^{th}$ individual. The predictor variable, $x_i$, corresponds to the numeric value assigned to each education level. The model can be described by the following equations:

$$
\begin{aligned}
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 +\beta_1x_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1) \\
\end{aligned}
$$
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

The model is executed in R[@citeR] using the rstanarm package [@rstanarm]. Default priors from rstanarm are utilized, set with a mean of zero and a conservative standard deviation, reflecting no strong prior beliefs about the magnitude of the education coefficients.

## Model justification

Given the economic theory and previous empirical findings, we anticipate that higher educational qualifications would correspond to higher hourly wages. This is encapsulated in the assumption that $\beta_i$ should have a positive effect on $y_i$. Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing empirical evidence to support or refine this hypothesis.

Mathematical notation is incorporated using LaTeX syntax, enabling precise formulation of our statistical model. For example, the average effect of education on hourly wage rate is denoted by $\beta$.

@@

The transformation from raw to analysis-ready data involved substantial cleaning, including filtering by geography (Canada), type of work (both full- and part-time), and specific wage type (average hourly rate). Moreover, we excluded broad or ambiguous categories like 'Some high school' and 'Total, all education levels' to ensure clarity in our analysis. The resulting 'Education_numeric' variable was constructed using a sorted factorization process, mapping each education level to a corresponding numeric value.

In presenting our findings, we will incorporate graphs that visualize the relationship between education and wage rates, providing an intuitive understanding of the data's characteristics. Tables summarizing the observations, along with discussions on the graphs, will be included to elucidate key trends and observations. Summary statistics and any notable relationships between the variables will also be featured, offering a detailed overview of the dataset. Should the volume of this information warrant it additional detail, these elements may be relegated to appendices to maintain the accessibility of the data section.
